iSEE: Interactive SummarizedExperiment Explorer

### 2018-05-14 manuscript/main.tex
Updated Makefile, added Mark to acknowledgements.
Added F1000Res editorial team changes.

\documentclass[10pt,a4paper,twocolumn]{article}
\usepackage{f1000_styles}
\usepackage{url}
\usepackage{listings}

@@

Most existing tools for interactive visualisation of biological data are designed for specific assays and analyses, e.g., pRoloc for proteomics \citep{gatto2014mass}, shinyMethyl for methylation \citep{fortin2014shinymethyl}, HTSvis for high-throughput screens \citep{scheeder2017htsvis}.
Opportunities for customisation are generally limited, making it difficult to re-use the same visualisation software for new technologies or experimental designs where different aspects of the data are of interest.
Moreover, standalone tools such as the Loupe Cell Browser from 10x Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}.
This complicates any coordinated use of these tools with a reproducible, transparent, and statistically rigorous analysis.\\

Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the Shiny framework \citep{chang2017shiny} and exploits data structures from the open-source Bioconductor project \citep{gentleman2004bioconductor}, specifically the SummarizedExperiment class.
iSEE allows users to simultaneously visualise multiple aspects of a given data set, including experimental data, metadata and analysis results.
Dynamic linking and point selection facilitate the flexible exploration of interactions between different data aspects.
Additional functionalities include code tracking, intelligent downsampling of large data sets, custom colour scale specification and tour construction.
We demonstrate the capabilities of iSEE by applying it to a diverse range of real data sets.

@@

\subsection*{Using a multi-panel layout}
All data aspects stored in a SummarizedExperiment can be simultaneously examined in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
The interface layout is built using the shinydashboard package \citep{chang2018shinydashboard}, using colour-coded panels to visualise each data aspect.
Individual panel types include:

\begin{itemize}
\item Column data plots, for visualising sample metadata stored in the colData slot of the SummarizedExperiment object.
\item Feature assay plots, for visualising expression values for a particular feature (e.g., gene) across samples from any assay in the SummarizedExperiment object.
\item Row statistics tables, to present the contents of the rowData slot for the SummarizedExperiment object.
\item Row data plots, for visualising feature metadata stored in the rowData slot of the SummarizedExperiment object.
\item Heatmaps, to visualise assay data for multiple features where samples are ordered by one or more colData fields.
\item Reduced dimension plots, which display any two dimensions from pre-computed dimensionality reduction results (e.g., from PCA or $t$-SNE).
These results are taken from the reducedDim slot if this is available in the object supplied to iSEE.
\end{itemize}

\begin{figure*}[t]
\includegraphics[width=\textwidth]{pics/Fig1.pdf}
\caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualises a different aspect of the data.
New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
Panel types are available to visualise sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualise experimental observations for multiple features; and feature-level metadata plots.
Panels of each type are colour-coded for ease of interpretation.
(B) Information can be transmitted between panels according to a user-specified scheme.
Here, the selection of feature $X$ in the row statistics table determines the y-axis of the feature assay plot, and colours the samples in the reduced dimension plot by the expression of $X$.
Selection of points in the reduced dimension plot (dotted blue line) also determines the samples that are shown in the column data (i.e., sample metadata) plot;
further selection of points in the column data plot determines the samples that are shown in the heatmap.
}
\label{fig:iSEE}
\end{figure*}

Each sample is represented as a point in column data, feature assay and reduced dimension plots.
Similarly, each feature is represented by a point in row data plots.
For these panel types, a scatter plot is automatically produced if the selected variables on the x- and y-axes are both continuous.
If exactly one variable is categorical, points are grouped by the categorical levels and a (vertical or horisontal) violin plot is produced with points scattered within each violin.
If both variables are categorical, a ``rectangle plot'' is produced where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of that combination.
Points are scattered randomly within each rectangle.
For ease of interpretation, the rectangle plot collapses to a mirrored bar plot when one of the categorical variables only has one level.

@@

By default, colour maps for categorical and continuous variables are taken from the ggplot2 \citep{wickham2009ggplot2} and viridis packages \citep{garnier2018viridis}, respectively.
However, iSEE also implements the ExperimentColorMap class, which allows users to specify arbitrary colour maps for particular variables.
Each colour map is a function that returns a vector of distinct colours of a specified length, and will be called whenever the associated variable is used for point colouring in a particular panel.
The returned colours will be mapped to factor levels for categorical variables, or used in colour interpolation for continuous variables.
For categorical variables, the function may also return a constant vector of named colours corresponding to the levels of a known factor.
Colour maps can be specified for individual variables; for all assays, all column data variables, or all row data variables (with different functions for continuous or categorical variables); or for all categorical or continuous variables.
This provides a convenient yet flexible mechanism for customisation of colouring schemes within the interface.

@@

\subsection*{Additional functionalities}
Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl \citep{zerbino2018ensembl} or Entrez \citep{ncbi2017entrez}. % do we need to cite them? at least ENSEMBL has updated citable papers
For large data sets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualisation.
Users can also include a bespoke step-by-step ``tour'' of their data set via the rintrojs package \cite{ganz2016rintrojs}, guiding the audience through an examination of the salient features in the data.

@@

Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE/}

@@

Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_PBMC4k/}

@@

Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_TCGA/}

@@

Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_cytof/}

@@

\section*{Data and software availability}
The iSEE package is available at \url{https://bioconductor.org/packages/iSEE}.
Source code of the development version of the package is available at \url{https://github.com/csoneson/iSEE}.
Code for the demonstrations and tours is available at \url{https://github.com/LTLA/iSEE2018}.

\section*{Author contributions}
All authors contributed conceptually and technically to the development and maintenance of large sections of the package code.
ATLL and FM developed the application and data infrastructure.
KRA and CS developed the plotting framework.
FM developed the code tracking and user interface.
KRA designed and developed the ExperimentColorMap class.
All authors contributed to testing the application on all data sets.
All authors wrote and approved the final manuscript.

\section*{Competing interests}
No competing interests were declared.

@@

\section*{Acknowledgements}
We thank the organisers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
We also thank members of the Bioconductor community for their helpful suggestions.
Finally, we thank John Marioni for helpful comments on the manuscript.

>>>>

\documentclass[10pt,a4paper,twocolumn]{article}
\usepackage{f1000_styles}
\usepackage{url}
\usepackage{listings}
\usepackage{hyperref}

@@

Most existing tools for interactive visualisation of biological data are designed for specific assays and analyses, e.g., \href{https://bioconductor.org/packages/release/bioc/html/pRoloc.html}{pRoloc} for proteomics \citep{gatto2014mass}, \href{http://bioconductor.org/packages/release/bioc/html/shinyMethyl.html}{shinyMethyl} for methylation \citep{fortin2014shinymethyl}, \href{http://htsvis.dkfz.de/HTSvis/}{HTSvis} for high-throughput screens \citep{scheeder2017htsvis}.
Opportunities for customisation are generally limited, making it difficult to re-use the same visualisation software for new technologies or experimental designs where different aspects of the data are of interest.
Moreover, standalone tools such as the \href{https://support.10xgenomics.com/single-cell-gene-expression/software/visualization/latest/what-is-loupe-cell-browser}{Loupe Cell Browser} from 10x Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}.
This complicates any coordinated use of these tools with a reproducible, transparent, and statistically rigorous analysis.\\

Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the \href{https://shiny.rstudio.com/}{Shiny framework}  \citep{chang2017shiny} and exploits data structures from the open-source \href{https://www.bioconductor.org/}{Bioconductor project}  \citep{gentleman2004bioconductor}, specifically the SummarizedExperiment class.
iSEE allows users to simultaneously visualise multiple aspects of a given data set, including experimental data, metadata and analysis results.
Dynamic linking and point selection facilitate the flexible exploration of interactions between different data aspects.
Additional functionalities include code tracking, intelligent downsampling of large data sets, custom colour scale specification and tour construction.
We demonstrate the capabilities of iSEE by applying it to a diverse range of real data sets.

\section*{Operation}
% * <research@f1000.com> 2018-05-14T09:51:45.749Z:
% 
% > Operation
% please complete this section regarding the minimal system requirements for your software tool and an overview of the work flow
% 
% ^.

@@

\subsection*{Using a multi-panel layout}
All data aspects stored in a SummarizedExperiment can be simultaneously examined in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
The interface layout is built using the \href{https://rstudio.github.io/shinydashboard/}{shinydashboard} package \citep{chang2018shinydashboard}, using colour-coded panels to visualise each data aspect.
Individual panel types include:

\begin{itemize}
\item Column data plots, for visualising sample metadata stored in the colData slot of the SummarizedExperiment object.
\item Feature assay plots, for visualising expression values for a particular feature (e.g. gene) across samples from any assay in the SummarizedExperiment object.
\item Row statistics tables, to present the contents of the rowData slot for the SummarizedExperiment object.
\item Row data plots, for visualising feature metadata stored in the rowData slot of the SummarizedExperiment object.
\item Heatmaps, to visualise assay data for multiple features where samples are ordered by one or more colData fields.
\item Reduced dimension plots, which display any two dimensions from pre-computed dimensionality reduction results (e.g., from PCA or $t$-SNE).
These results are taken from the reducedDim slot if this is available in the object supplied to iSEE.
\end{itemize}

\begin{figure*}[t]
\includegraphics[width=\textwidth]{Fig1.pdf}
\caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualises a different aspect of the data.
New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
Panel types are available to visualise sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualise experimental observations for multiple features; and feature-level metadata plots.
Panels of each type are colour-coded for ease of interpretation.
(B) Information can be transmitted between panels according to a user-specified scheme.
Here, the selection of feature $X$ in the row statistics table determines the y-axis of the feature assay plot, and colours the samples in the reduced dimension plot by the expression of $X$.
Selection of points in the reduced dimension plot (dotted blue line) also determines the samples that are shown in the column data (i.e., sample metadata) plot;
further selection of points in the column data plot determines the samples that are shown in the heatmap.
}
\label{fig:iSEE}
\end{figure*}

Each sample is represented as a point in column data, feature assay and reduced dimension plots.
Similarly, each feature is represented by a point in row data plots.
For these panel types, a scatter plot is automatically produced if the selected variables on the x- and y-axes are both continuous.
If exactly one variable is categorical, points are grouped by the categorical levels and a (vertical or horizontal) violin plot is produced with points scattered within each violin.
If both variables are categorical, a ``rectangle plot'' is produced where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of that combination.
Points are scattered randomly within each rectangle.
For ease of interpretation, the rectangle plot collapses to a mirrored bar plot when one of the categorical variables only has one level.

@@

By default, colour maps for categorical and continuous variables are taken from the \href{http://ggplot2.tidyverse.org/}{ggplot2}  \citep{wickham2009ggplot2} and \href{https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html}{viridis} packages \citep{garnier2018viridis}, respectively.
However, iSEE also implements the ExperimentColorMap class, which allows users to specify arbitrary colour maps for particular variables.
Each colour map is a function that returns a vector of distinct colours of a specified length, and will be called whenever the associated variable is used for point colouring in a particular panel.
The returned colours will be mapped to factor levels for categorical variables, or used in colour interpolation for continuous variables.
For categorical variables, the function may also return a constant vector of named colours corresponding to the levels of a known factor.
Colour maps can be specified for individual variables; for all assays, all column data variables, or all row data variables (with different functions for continuous or categorical variables); or for all categorical or continuous variables.
This provides a convenient yet flexible mechanism for customisation of colouring schemes within the interface.

@@

\subsection*{Additional functionalities}
Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as \href{https://www.ensembl.org/index.html}{Ensembl} \citep{zerbino2018ensembl} or  \href{https://www.ncbi.nlm.nih.gov/Class/MLACourse/Original8Hour/Entrez/}{Entrez} \citep{ncbi2017entrez}. % do we need to cite them? at least ENSEMBL has updated citable papers
For large data sets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualisation.
Users can also include a bespoke step-by-step ``tour'' of their data set via the \href{https://introjs.com/}{rintrojs} package \cite{ganz2016rintrojs}, guiding the audience through an examination of the salient features in the data.

@@

An interactive tour of this use case can be viewed \href{http://shiny.imbei.uni-mainz.de:3838/iSEE/}{here}

@@

An interactive tour of this use case can be viewed \href{http://shiny.imbei.uni-mainz.de:3838/iSEE_PBMC4k/}{here}

@@

An interactive tour of this use case can be viewed \href{http://shiny.imbei.uni-mainz.de:3838/iSEE_TCGA/}{here}

@@

An interactive tour of this use case can be viewed \href{http://shiny.imbei.uni-mainz.de:3838/iSEE_cytof/}{here}

@@

\section*{Software availability}
% * <research@f1000.com> 2018-05-14T10:15:33.183Z:
% 
% > Data and software availability
% please provide an archived version of you code and the licence under which it can be used, see email
% 
% ^.
The iSEE package is available at \url{https://bioconductor.org/packages/iSEE}.
Source code of the development version of the package is available at \url{https://github.com/csoneson/iSEE}.
Code for the demonstrations and tours is available at \url{https://github.com/LTLA/iSEE2018}.
% * <research@f1000.com> 2018-05-14T10:10:51.545Z:
% 
% The author contribution section has been deleted, since CRediT was filled in on submission and will be used for the Author Contributions section.
% 
% ^.
\section*{Data availability }

Data used in the described use cases is available from the following articles: \\
\url{http://doi.org/10.1038/nn.4216}\citep{tasic2016allen} \\
\url{http://doi.org/10.1038/ncomms14049}\citep{zheng2017massively} \\
\url{https://doi.org/10.1093/bioinformatics/btv377}\citep{piccolo2015TCGA} \\
\url{https://doi.org/10.1038/nbt.2317}\citep{bodenmiller2012cytof} \\
\section*{Competing interests}
No competing interests were declared.

@@

\section*{Acknowledgements}
We thank the organisers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
We also thank members of the Bioconductor community for their helpful suggestions.
Finally, we thank John Marioni for helpful comments on the manuscript.
% * <research@f1000.com> 2018-05-14T10:11:46.639Z:
% 
% > John Marioni 
% please confirm you have permission to name them in your manuscript 
% 
% ^.
